### 2025-07-14 manuscript/part2.tex
custom scope provider

    return exports;
  }
}
\end{minted}
\subsection*{Scope provider}
import

>>>>

    return exports;
  }
}
\end{minted}
\captionof{listing}{Custom ScopeComputation service}

\subsection*{Scope provider}

The custom scope provider focuses specifically on two kinds of nodes - function calls (\code{FuncCall}) and variable references (\code{LVal}) - since
these are the only AST nodes in the MiniProb grammar that establish referential links to other symbols. By subclassing the base \code{ScopeProvider} and overriding
its \code{getScope} method, the provider can dispatch to specialized logic for these two cases while delegating all other node types back to the default implementation.
This fallback invocation of \code{super.getScope()} outside of the \code{FuncCall} and \code{LVal} branches ensures that any other or future nodes will receive a fully defined and well-formed scope, preserving completeness of the overall resolution process.

Specialized handling is required because the language supports an import mechanism that selectively brings in only top-level definitions from other files, while excluding
all other symbols, which are exported by default into a workspace-wide scope governed by the \code{IndexManager}. In the \code{LVal} branch, the provider correctly resolves
multiple identifiers declared in a single declarations statement (for example, \code{u8 i1, i2, i3}) distinguishing between block-scope and global variables, as well as including function parameters as valid resolutions.
In the \code{FuncCall} branch, default behavior is overridden to prevent 'ghost' functions - placeholder symbols automatically invented for unresolved identifiers - and to include in-flight local declarations that appear before their closing delimiters in live editing scenarios (a solution which inspects raw \code{CST} text and may require future revision).

Because both \code{FuncCall} and \code{LVal} reference the same identifier token pattern, \code{getScope} is invoked for both branches on the same node.
If distinguishing syntactic context - such as surrounding parentheses for calls or the presence of an equals sign for assignments - is present, the provider resolves only once aware of what path to choose.
\\

The helper method below builds a resolution scope for symbols imported from other files. It first transforms each import entry into an absolute URI by joining the current
document's directory path with the relative import path. Next, it invokes the \code{IndexManager} to collect all symbol descriptions of the requested type across the entire
workspace, filtering to only those defined in the imported files. Notably, only symbols of already parsed document are managed by the \code{IndexManager}.
To improve performance, the retrieved descriptions are stored in a cache keyed by the document URI and symbol kind, and wrapped in a MapScope for efficient lookup. 
\begin{minted}[bgcolor=lightgray, frame=lines, breaklines, linenos]{ts}
  private getImportedScope(
    fileImports: FileImport[],
    currentUri: URI,
    targetNodeType: string
  ): Stream<AstNodeDescription> {
    const importUris = fileImports.map((f) => {
      const filePath = posix.join(dirname(currentUri.path), f.file);
      return currentUri.with({ path: filePath }).toString();
    });

    const importKey = 'imported-';
    if (targetNodeType === Func) {
      const importedFuncDescriptions = this.descriptionCache.get(
        currentUri,
        importKey + targetNodeType,
        () => new MapScope(this.indexManager.allElements(Func, new Set<string>(importUris)))
      );
      return importedFuncDescriptions.getAllElements();
    } 
    .
    .
  }
\end{minted}
\captionof{listing}{Inclusion of imported file's top-level scope}
\vspace{0.3em}

In this snippet, the overridden \code{getScope} method first confirms that the current resolution request pertains to any \code{reference} context by checking
\code{context.property} and container presence (lines 1-3), then locates the enclosing program node. Upon detecting a function-call node (line 11), it filters the program's
function list to remove placeholder or 'ghost' entries (line 7). A local cache - persistent throughout the whole linking phase - is
then used to retrieve or populate descriptors for these real functions (lines 11-22), encapsulating them within Langium's offered scope structure. If file-level imports are present
(lines 8, 26-30), a helper routine leverages the global \code{IndexManager} to fetch additional symbols from imported modules, yielding a secondary stream of descriptors.
The in-file and imported descriptors, together with any local declarations, are merged into a composite scope (lines 32-34) that is returned for code completion and reference
linking; all other situations fall back to \code{super.getScope()} (lines 38), preserving the default resolution behavior.

\begin{minted}[bgcolor=lightgray, frame=lines, breaklines, linenos]{ts}
  override getScope(context: ReferenceInfo): Scope {
    const container = context.container;
    if (context.property === 'ref' && container) {
      const program = AstUtils.getContainerOfType(container, isProgram)!;

      // filter Func for body -> only real Func and not ghost Reference(=current input)
      const programFunctions = program.functions.filter(this.isRealFunc);
      const includeFileImports = program.fileImports && program.fileImports.length > 0;

      var importedDescriptions: Stream<AstNodeDescription> = stream();
      if (isFuncCall(container)) {
        const descriptions = this.descriptionCache
          .get(
            AstUtils.getDocument(container).uri,
            Func,
            () =>
              new MapScope(
                programFunctions.map((func) =>
                  this.astNodeDescriptionProvider.createDescription(func, func.name)
                )
              )
          ).getAllElements();
        .
        .      
        //check for imported functions
        if (includeFileImports) {
          const document = AstUtils.getDocument(container);
          const uri = document.uri;
          importedDescriptions = this.getImportedScope(program.fileImports, uri, Func);
        }

        return new MapScope(
          stream(descriptions, importedDescriptions, localDeclarationsDescriptions)
        );
      } 
      .
      .
    return super.getScope(context);
}}
\end{minted}
